From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-20 Pi+/paper/finite.tex
Revise Coxeter section

We consider the generators for the Coxeter relations on $\List[\Fin[n]]$ as a
rewriting system. Throwing out reflexivity, and symmetry, we observe that the
right hand sides of the relations are strictly smaller than the left hand sides,
in terms of the lexicographical ordering on words in $\Fin[n]$. By directing the
relation from left to right, we would be able to prove a termination property
out of the box.

However, it is not at all obvious that the system would be confluent.
There are many possible critical pairs, such as
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXGJyYWlkIiwxXV0=
\[\begin{tikzcd}
    && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
    {\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\braid"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]
or
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFs0LDEsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiXFxicmFpZCIsMV0sWzAsMiwiXFxjYW5jZWwiLDFdXQ==
\[\begin{tikzcd}
    && {\tau_2\tau_1\tau_2\tau_2} \\
    {\tau_1\tau_2\tau_1\tau_2} &&&& {\tau_2\tau_1}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\cancel"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]
Although in the cases above, critical pairs resolve, there are some examples where they do not. One of the simplest ones is the following:
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMSwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFs0LDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXHN3YXAiLDFdXQ==
\[\begin{tikzcd}
    && {\tau_3\tau_2\tau_3\tau_1} \\
    {\tau_2\tau_3\tau_2\tau_1} &&&& {\tau_3\tau_2\tau_1\tau_3}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\swap"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]

Because of this counterexample, the equations have to be fixed. We thus propose a new presentation, keeping the
$\cancel$ and $\swap$ constructors, but replace $\braid$ with a more general $\longbraid$:

Diagram of long-braid:

@@

@@

@@

\review{%
  In return for increased complexity of the equations, it satisfies the diamond property (if we direct the equation).
  Thus, together with the (preserved) termination, it creates a confluent rewriting system. Proof: All the critical
  pairs resolve, in the appendix. Definition: A normal form is a list that cannot be reduced any further.%
}

Now, we state and prove that $(\List[\Fin[n]],\longcox*)$ is a confluent and strongly normalising rewriting system.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\longcox*$ is (locally) confluent. For every span $\longcoxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is confluent. For every extended span $\longcoxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
    \item $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
  \end{enumerate}
\end{proposition}

@@

19. However, the Long Coxeter relations are unwieldly and have no
connectino to the broader scope of computational group theory - while the
usual Coxeter presentation is to-go choice in this area (cite all these
papers in Coq, this guy from Sweden(?) etc.).
Thus, we prove:
Proposition: Long Coxeter is equivalent to normal Coxeter.

\todo{Long form of Coxeter relations and their equivalence.}

19a. (Note: here we try to convince the reader about reusability badge) So,
the Long Coxeter can be regarded as a trick to get the proof to pass - but
the external user-facing interface is still that of the usual Coxeter
presentation.

@@

We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}. \vc{Not completely
  formalised. Is this with Coxeter or LongCoxeter?}

>>>>

To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. The $\cox*$ relation as defined, does not produce a
well-behaved rewriting system due to the symmetric closure. Throwing out reflexivity and symmetry, we observe that the
right hand sides of the relations are strictly smaller than the left hand sides, in terms of the lexicographical
ordering on words in $\Fin[n]$. By directing the relation from left to right, we would be able to prove a termination
property out of the box.

A well-behaved rewriting system should also be confluent, however, it is not at all obvious from this definition. Due to
overlapping reduction rules, there are many possible critical pairs, such as

\[
  \begin{array}{lcr}
    \gspan[\braid][\braid]{\tau_2\tau_1\tau_2\tau_1\tau_2}{\tau_1\tau_2\tau_1\tau_1\tau_2}{\tau_2\tau_1\tau_1\tau_2\tau_1}
     &
    \text{or}
     &
    \gspan[\braid][\cancel]{\tau_2\tau_1\tau_2\tau_2}{\tau_1\tau_2\tau_1\tau_2}{\tau_2\tau_1}
  \end{array}
\]

For confluence, all critical pairs should converge. The pairs above can be resolved, but there are many examples where
they do not. The following is a simple example.

\[
  \gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}
\]

Because of this counterexample, the relations have to be changed. We propose a long form of the relations as a new
presentation, where we allow $\cancel$ and $\swap$ reductions anywhere inside the list, called $\longcancel$ and
$\longswap$, but replace $\braid$ with a more general $\longbraid$ relation.

@@

The directed $\longcox$ relation is defined formally with the following generators.

@@

Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity. The
transitive extension is only performed on the right.

@@

Despite the increased complexity of the relations, the rewriting system $(\List[\Fin[n]],\longcox*)$ can be shown to
satisfy confluence, that is, the Church-Rosser (diamond) property. Further, we can show that this system is strongly
normalising, by producing a unique normal form that doesn't reduce any further. We follow the terminology
of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\longcox$ is (locally) confluent. For every span $\coxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
  \end{enumerate}
\end{proposition}

Another property we're interested in is well-foundedness of the relation $\longcox*$.
\todo{lexicographic order, accessibility predicate and well-foundedness}

In the theory of group presentations, the Coxeter relations are generally presented in their short form~\todo{citation},
but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}. The long form of the
Coxeter relations are unwieldy and difficult to prove properties about by induction, and have little connection to the
broader scope of computational group theory. However, we can prove that the closures of the two forms of the relations
are equivalent, and work up to this equivalence!

\begin{proposition}
  $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
\end{proposition}

@@

@@
